---
title: "Vis: Histograms"
author: Zach del Rosario
date: 2020-05-22
output: github_document
time: 30
reading: 20
---

# Vis: Histograms

*Purpose*: *Histograms* are a key tool for EDA. In this exercise we'll get a little more practice constructing and interpreting histograms and densities.

*Reading*: [Histograms](https://rstudio.cloud/learn/primers/3.3) *Topics*: (All topics) *Reading Time*: \~20 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### **q1** Using the graphs generated in the chunks `q1-vis1` and `q1-vis2` below, answer:

-   Which `class` has the most vehicles?
    -   It seems like the suv class has the most cars, but we need to add together across all bins of city miles per gallon. These plots are not good for answering this question.
-   Which `class` has the broadest distribution of `cty` values?
    -   Pickups have the broadest distribution of cty values.
-   Which graph---`vis1` or `vis2`---best helps you answer each question?
    -   Neither particularly helps with the first one, the smoothed graph helps more with the second question.

```{r q1-vis1}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_freqpoly(bins = 10)
```

```{r q1-vis2}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density()
```

In the previous exercise, we learned how to *facet* a graph. Let's use that part of the grammar of graphics to clean up the graph above.

### **q2** Modify `q1-vis2` to use a `facet_wrap()` on the `class`. "Free" the vertical axis with the `scales` keyword to allow for a different y scale in each facet.

```{r q2-task}
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density() +
  facet_wrap( ~class, scales = "free")
```

In the reading, we learned that the "most important thing" to keep in mind with `geom_histogram()` and `geom_freqpoly()` is to *explore different binwidths*. We'll explore this idea in the next question.

### **q3** Analyze the following graph; make sure to test different binwidths. What patterns do you see? Which patterns remain as you change the binwidth?

```{r q3-task}
## TODO: Run this chunk; play with differnet bin widths
diamonds %>%
  filter(carat < 1.1) %>%

  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.1, boundary = 0.005) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.1)

  )
diamonds %>%
  filter(carat < 1.1) %>%

  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.01, boundary = 0.005) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.1)

  )
diamonds %>%
  filter(carat < 1.1) %>%

  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.001, boundary = 0.005) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.1)

  )
```

**Observations**

-   Sometimes larger binwidths help with zooming out and understanding the big picture, but can sometimes hide peaks that only show up with smaller binwidths.

-   With smaller binwidths peaks at 0,3, 0.5, 0.7, 0.9, and 1 show up. Possibly these are commonly sold carats of diamonds.

-   Using binwidths finer than 0.01 does not yield any new peaks, because the carat data does not have as fine resolution.

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis02-histograms-assignment.Rmd).
