---
title: "Vis: Lines"
author: Zach del Rosario
date: 2020-06-04
output: github_document
time: 15
reading: 30
editor_options: 
  markdown: 
    wrap: 72
---

# Vis: Lines

*Purpose*: *Line plots* are a key tool for EDA. In contrast with a
scatterplot, a line plot assumes the data have a *function* relation.
This can create an issue if we try to plot data that do not satisfy our
assumptions. In this exercise, we'll practice some best-practices for
constructing line plots.

*Reading*: [Line plots](https://rstudio.cloud/learn/primers/3.6)
*Topics*: Welcome, Line graphs, Similar geoms (skip Maps) *Reading
Time*: \~30 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
library(gapminder)

```

### **q1** The following graph doesn't work as its author intended. Based on what we learned in the reading, fix the following code.

```{r q1-task}
## TODO: Fix the following code
gapminder %>%
  filter(continent == "Asia") %>%
  ggplot(aes(year, lifeExp, color = country)) +
  geom_line()
```

### **q2** A line plot makes *a certain assumption* about the underlying data. What assumption is this? How does that assumption relate to the following graph? Put differently, why is the use of `geom_line` a bad idea for the following dataset?

```{r q2-vis}
## TODO: No need to edit; just answer the questions
mpg %>%
  ggplot(aes(displ, hwy)) +
  geom_line()

```

**Observations**: - What does a line plot assume about the data? - How
does the `mpg` dataset violate this assumption?

-   The line geom assumes that there is a single value for each x value,
    which causes straight vertical lines when it has to jump up and down
    to the different hwy mpg for each car in the dataset. We can fix
    this by grouping by car.

### **q3** The following graph shows both the raw data and a smoothed version. Describe the trends that you can see in the different curves.

```{r q3-task}
## TODO: No need to edit; just interpret the graph
economics %>%
  ggplot(aes(date, unemploy)) +
  geom_line(aes(color = "Raw")) +
  geom_smooth(aes(color = "Smoothed"), se = FALSE) +
  scale_color_discrete(name = "Source")
```

**Observations**: - What can you tell based on the `Raw` data? - What
can you tell based on the `Smoothed` data?

-   The raw data shows peaks that are hidden in the smoothed data.

-   The smoothed data is a lot easier to visualize the general trend of
    the data.

    -   The general data trends towards a higher number of unemployed
        people.

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit
ticket survey**, [linked
here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis05-lines-assignment.Rmd).
