---
title: "Gapminder"
author: "Drew Pang"
date: 2023-02-26
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
glimpse(gapminder)
```

**Observations**:

-   Variables:
    -   country

    -   continent

    -   year

    -   lifeExp

    -   pop

    -   gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(pull(gapminder, year))
year_min <- min(pull(gapminder, year))
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = gdpPercap, y = continent)) + 
  geom_boxplot() + 
  coord_flip()

gapminder %>%
  filter(year == year_min & gdpPercap < 30000) %>%
  ggplot(aes(x = gdpPercap, y = continent)) + 
  geom_boxplot() + 
  coord_flip()
```

**Observations**:

-   Oceania has the highest median `gdpPercap`, but closest quantiles, suggesting that there is a small sample size.

-   Median `gdpPercap` from highest to lower for continents is Oceania, Europe, Americas, Asia, Africa.

-   Europe has the greatest range without the very large outlier in Asia, and Oceania has the smallest range.

-   Europe has the largest box - or the largest range between the 25th and 75th percentiles.

**Difficulties & Approaches**:

-   My first thought was to create a boxplot, because we are plotting a continuous variable against a discrete variable. However, upon creating a simple boxplot, one of the outliers extends the scale of gdpPercap enough that it squishes all of the boxes towards the bottom quarter of the plot.
-   I simply added another filter to filter out gdpPercap values above 30000. Not sure what the correct way to go about removing outliers for the sake of graph visibility, but it made reading the box plot a lot easier.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min) %>%
  arrange(desc(gdpPercap))
```

**Observations**:

-   Identify the outlier countries from q2
    -   There are no outliers with lower than median GDP per capita, so I sorted the data in ascending order based on `gdpPercap`.
    -   The top three are Kuwait, Switzerland, and the United States.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min | year == year_max) %>%
  ggplot(aes(x = gdpPercap, y = continent)) + 
  geom_boxplot() + 
  coord_flip() + 
  facet_wrap(vars(year)) + 
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "Switzerland", "United States")),
    mapping = aes(color = country),
    size = 2
  )
```

**Observations**:

-   All three outliers from 1952 are still all above the median for their respective continents.
-   Kuwait is no longer as far away from the median GDP per capita for Asia as is was in 1952 in 2007, but still appears to be one of the highest, if not the highest in Asia.
-   The United States appears to have the highest GDP per capita in the Americas.
-   It's possible World War 2 and oil had to do with Kuwait's very large `gdpPercap` in 1952. More research is necessary to determine the root cause.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  filter(year == year_min | year == year_max) %>%
  ggplot(aes(x = lifeExp, y = continent)) + 
  geom_boxplot() + 
  coord_flip() + 
  facet_wrap(vars(year)) + 
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "Switzerland", "United States")),
    mapping = aes(color = country),
    size = 2
  )
```

-   Across the board, life expectancy has gone up in all continents.
-   Despite Kuwait's very large GDP per capita, their life expectancy is not the highest in the world, let alone just in Asia.
-   The Americas have the largest range for life expectancy in 1952, while Africa has the largest range in 2007. It is possible that there is a large development disparity in Africa.

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  filter(year == year_max) %>%
  arrange(desc(lifeExp))

gapminder %>%
  filter(country == "Japan" | country == "Hong Kong, (China)" | country == "Iceland" | country == "Switzerland" | country == "United States") %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) + 
  geom_line()
```

-   I first filtered for the max year observations, then looked at the top countries for life expectancy. I pulled the top three along with the United States as a refrence.
-   Life expectancy in these four countries goes up a noticeable amount. I though it would be a lot less, but a 10 year life expectancy increase in all countries is quite an improvement in just 50 years.

```{r q5-task3}
## TASK: Your third graph
gapminder %>%
  filter((year == year_min | year == year_max) & pop < 5e8) %>%
  ggplot(aes(x = pop, y = continent)) + 
  geom_boxplot() + 
  coord_flip() + 
  facet_wrap(vars(year)) + 
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "Switzerland", "United States")),
    mapping = aes(color = country),
    size = 2
  )
```

-   In general, the world population goes up. Asia in particular has a much larger range between the 25th and 75th percentiles.
-   The United States remained
